---
title: "TCGA REMP Retroelement Analysis"
author: "Chinaza Nnam"
date: "2026-02-27"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

## Overview

This script performs repetitive element (RE) methylation analysis on TCGA-KIRC data using REMP
predictions. It covers:

1. REMP annotation and prediction for Alu, L1, and LTR families
2. Trimming, aggregation, and KNN imputation of predicted beta/M values
3. Multi-omics integrative clustering (MOVICS/NEMO) to define RE methylation subtypes
4. Phenotype preparation and HiTIMED immune deconvolution
5. Differential methylation analysis (DMA) across subtypes
6. Genomic annotation of significant hits

**Required input objects (must be loaded before running):**
- `betas` — QC'ed CpG beta matrix (samples × CpGs) from TCGA-KIRC
- `pheno` — phenotype table with columns: `cases`, `basename`, `submitter_id`,
  `days_to_death`, `days_to_last_follow_up`, `vital_status`, `ajcc_pathologic_stage`,
  `tumor_grade`, `age_at_index`, `gender`, `race`

**Outputs flow into:** `02_Survival_RE_Analysis.Rmd` and
`03_Hypoxia_ImmuneResponse_Signatures_RE_Analysis.Rmd`

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "results/figures/"   # all figures saved here automatically
)

# Create required output directories if they don't already exist
dir.create("data",            showWarnings = FALSE, recursive = TRUE)
dir.create("results",         showWarnings = FALSE, recursive = TRUE)
dir.create("results/figures", showWarnings = FALSE, recursive = TRUE)
```

---

## Load Packages

```{r packages}
# Data wrangling
library(dplyr)
library(tidyr)
library(ggplot2)

# REMP methylation prediction and imputation
library(REMP)
library(impute)

# Multi-omics integrative clustering
library(MOVICS)

# Survival analysis
library(survival)
library(survminer)
library(broom)

# Differential methylation (EWAS)
library(limma)

# Genomic range overlaps for annotation
library(GenomicRanges)

# Summary tables
library(tableone)

# Volcano plots — uncomment if EnhancedVolcano is installed
# library(EnhancedVolcano)

# Additional optional packages (uncomment as needed):
# library(ggpubr)
# library(rstatix)
# library(GSVA)
# library(BiocParallel)
# library(VIM)
# library(matrixStats)
```

---

## Helper Functions

All custom utility functions are defined here so they can be reused throughout the script.

```{r helper-functions}
# --- Path helpers ---
# Prepend "data/" or "results/" to a filename for consistent file I/O
p_data <- function(x) file.path("data", x)
p_res  <- function(x) file.path("results", x)

# --- Column name cleaner ---
# Strips a leading "X" that R sometimes adds to numeric column names (e.g. "X123" → "123")
removeX <- function(mat) {
  colnames(mat) <- gsub("^X", "", colnames(mat))
  mat
}

# --- KNN imputation wrapper ---
# Runs impute.knn() with consistent defaults across all matrices in this project
# rowmax/colmax: maximum proportion of missing values allowed per row/column before removal
impute_knn_mat <- function(mat, k = 10, rowmax = 0.5, colmax = 0.8) {
  out <- impute.knn(as.matrix(mat), k = k, rowmax = rowmax, colmax = colmax)
  out$data
}

# --- Logit (beta → M-value) transform ---
# Converts beta values (0–1) to M-values (-Inf to +Inf) for linear modelling
# eps clips values away from 0 and 1 to avoid log(0) = -Inf
logit_beta <- function(beta, eps = 1e-6) {
  beta <- pmin(pmax(beta, eps), 1 - eps)
  log2(beta / (1 - beta))
}

# --- Extract hazard ratio from Cox model for a specific NEMO cluster ---
# Returns HR, 95% CI, and p-value as a tidy data frame
extract_nemo_hr <- function(model, nemo_prefix = "NEMO", level_pattern = "Active") {
  sm  <- summary(model)
  rn  <- rownames(sm$coefficients)
  nemo_rows <- grep(paste0("^", nemo_prefix), rn, value = TRUE)
  target    <- grep(level_pattern, nemo_rows, value = TRUE)
  if (length(target) == 0) stop("Requested NEMO level not found in model coefficients.")
  coef_row <- sm$coefficients[target[1], ]
  ci_row   <- sm$conf.int[target[1], ]
  data.frame(
    term  = target[1],
    HR    = unname(ci_row["exp(coef)"]),
    lower = unname(ci_row["lower .95"]),
    upper = unname(ci_row["upper .95"]),
    pval  = unname(coef_row["Pr(>|z|)"]),
    row.names = NULL
  )
}

# --- Tidy Cox model output for forest plots ---
# Exponentiates coefficients (→ HR) and adds a label column for multi-model comparisons
tidy_models <- function(model, label) {
  broom::tidy(model, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(Model = label)
}

# --- Genomic overlap join ---
# Joins an annotation data frame to an RE position data frame based on genomic overlap
# Returns one row per overlapping pair with columns from both data frames
overlap_join_annotations <- function(annot_df, re_df) {
  gr_annot <- GRanges(
    seqnames = annot_df$seqnames,
    ranges   = IRanges(start = annot_df$start, end = annot_df$end),
    strand   = if ("strand" %in% names(annot_df)) annot_df$strand else "*"
  )
  mcols(gr_annot) <- annot_df %>% select(-seqnames, -start, -end, -strand)

  gr_re <- GRanges(
    seqnames = re_df$seqnames,
    ranges   = IRanges(start = re_df$start, end = re_df$end),
    strand   = if ("strand" %in% names(re_df)) re_df$strand else "*"
  )
  mcols(gr_re) <- re_df %>% select(-seqnames, -start, -end, -strand)

  ov <- findOverlaps(gr_re, gr_annot, ignore.strand = TRUE)
  bind_cols(
    as.data.frame(gr_re[queryHits(ov)]),
    as.data.frame(mcols(gr_annot)[subjectHits(ov), ])
  )
}
```

---

## Check Required Inputs

```{r check-inputs}
# Stop early with a helpful message if required objects are not in the environment.
# Load betas and pheno from your data/ folder before running this notebook.
required <- c("betas", "pheno")
missing  <- required[!vapply(required, exists, logical(1))]
if (length(missing) > 0) {
  stop(
    "Missing required objects: ", paste(missing, collapse = ", "),
    ". Please load them into your R environment before knitting."
  )
}
```

---

## Step 1: Build REMP Annotations

REMP requires pre-built annotation parcels for each RE family. These are cached to `data/` so
they only need to be downloaded once — subsequent runs load from disk.

```{r remp-annotations}
alu_rda <- p_data("remparcelAlu_450k_hg38.rda")
l1_rda  <- p_data("remparcelL1_450k_hg38.rda")
ltr_rda <- p_data("remparcelLTR_450k_hg38.rda")

# Alu
if (!file.exists(alu_rda)) {
  remparcelAlu <- initREMP(arrayType = "450k", REtype = "Alu",
                           annotation.source = "AH", genome = "hg38", ncore = 5)
  save(remparcelAlu, file = alu_rda)
} else {
  load(alu_rda)
}

# L1
if (!file.exists(l1_rda)) {
  remparcelL1 <- initREMP(arrayType = "450k", REtype = "L1",
                          annotation.source = "AH", genome = "hg38", ncore = 5)
  save(remparcelL1, file = l1_rda)
} else {
  load(l1_rda)
}

# LTR (ERV family)
if (!file.exists(ltr_rda)) {
  remparcelLTR <- initREMP(arrayType = "450k", REtype = "ERV",
                           annotation.source = "AH", genome = "hg38", ncore = 5)
  save(remparcelLTR, file = ltr_rda)
} else {
  load(ltr_rda)
}
```

---

## Step 2: Run REMP Predictions

REMP predicts methylation levels at RE loci using nearby CpGs from the array as predictors.
Results are cached to `data/` for reproducibility.
betas = your hiTAIC QC'ed and ENmix Preprocessed and Q'Ced beta matrix for tcga tumors
```{r remp-predictions}
alu_out <- p_data("remp.resAlu_tcga.rda")
l1_out  <- p_data("remp.resL1_tcga.rda")
ltr_out <- p_data("remp.resLTR_tcga.rda")

if (!file.exists(alu_out)) {
  remp.resAlu <- remp(betas, REtype = "Alu", parcel = remparcelAlu, ncore = 20, seed = 777)
  save(remp.resAlu, file = alu_out)
} else load(alu_out)

if (!file.exists(l1_out)) {
  remp.resL1 <- remp(betas, REtype = "L1", parcel = remparcelL1, ncore = 20, seed = 777)
  save(remp.resL1, file = l1_out)
} else load(l1_out)

if (!file.exists(ltr_out)) {
  remp.resLTR <- remp(betas, REtype = "ERV", parcel = remparcelLTR, ncore = 20, seed = 777)
  save(remp.resLTR, file = ltr_out)
} else load(ltr_out)
```

---

## Step 3: Trim, Aggregate, Extract Matrices, and Impute

- **Trim:** removes low-confidence predictions (reliability score threshold = 1.7) and loci with
  >20% missing values
- **Aggregate:** collapses predictions to RE locus level, requiring ≥2 CpGs per locus
- **Impute:** KNN imputation fills remaining missing values

```{r trim-aggregate-impute}
# Trim low-quality predictions
remp.resAlu2 <- rempTrim(remp.resAlu, threshold = 1.7, missingRate = 0.2)
remp.resL12  <- rempTrim(remp.resL1,  threshold = 1.7, missingRate = 0.2)
remp.resLTR2 <- rempTrim(remp.resLTR, threshold = 1.7, missingRate = 0.2)

# Aggregate to RE locus level (minimum 2 CpGs per locus)
remp.resAluag <- rempAggregate(remp.resAlu2, NCpG = 2)
remp.resL1ag  <- rempAggregate(remp.resL12,  NCpG = 2)
remp.resLTRag <- rempAggregate(remp.resLTR2, NCpG = 2)

# Extract beta (methylation proportion) and M-value (logit-transformed) matrices
Alu_beta <- as.data.frame(remp.resAluag@assays@data$rempB)
L1_beta  <- as.data.frame(remp.resL1ag@assays@data$rempB)
LTR_beta <- as.data.frame(remp.resLTRag@assays@data$rempB)

Alu_M <- as.data.frame(remp.resAluag@assays@data$rempM)
L1_M  <- as.data.frame(remp.resL1ag@assays@data$rempM)
LTR_M <- as.data.frame(remp.resLTRag@assays@data$rempM)

# KNN imputation (k=10) for remaining missing values in beta and M matrices
Alu_beta2 <- impute_knn_mat(Alu_beta)
L1_beta2  <- impute_knn_mat(L1_beta)
LTR_beta2 <- impute_knn_mat(LTR_beta)

Alu_M2 <- impute_knn_mat(Alu_M)
L1_M2  <- impute_knn_mat(L1_M)
LTR_M2 <- impute_knn_mat(LTR_M)
```

---

## Step 4: Select Elite Features and Run MOVICS Clustering

The top 1,000 most variable loci per RE family (by MAD) are selected as features for clustering.
Six algorithms are run; downstream analyses use NEMO as the primary assignment.

```{r movics-clustering}
# Select top 1,000 most variable loci per family using Median Absolute Deviation (MAD)
Alu_1k <- getElites(dat = Alu_beta2, method = "mad", elite.num = 1000)
L1_1k  <- getElites(dat = L1_beta2,  method = "mad", elite.num = 1000)
LTR_1k <- getElites(dat = LTR_beta2, method = "mad", elite.num = 1000)

# Assemble multi-omics input list
mo.data <- list(
  Alu = Alu_1k$elite.dat,
  L1  = L1_1k$elite.dat,
  LTR = LTR_1k$elite.dat
)

# Determine optimal cluster number (k=3 was selected based on this output)
optk.RE <- getClustNum(
  data        = mo.data,
  try.N.clust = 2:8,
  fig.name    = "CLUSTER_NUMBER_TCGA_KIRC_RE"
)

# Run 6 multi-omics clustering algorithms with k=3
# All data types are continuous methylation values (gaussian distribution assumed)
moic.res.list <- getMOIC(
  data        = mo.data,
  methodslist = list("SNF", "NEMO", "COCA", "ConsensusClustering", "IntNMF", "MoCluster"),
  N.clust     = 3,
  type        = c("gaussian", "gaussian", "gaussian")
)

# Uncomment to save clustering results for use in downstream scripts:
# save(moic.res.list, file = p_data("moic.res.list.rda"))
```

---

## Step 5: Consensus Clustering and Heatmap

Consensus clustering aggregates results across all 6 algorithms to identify the most stable
sample assignments. The heatmap visualises cluster structure after logit transformation and
standardisation.

```{r consensus-heatmap}
# Consensus across all 6 algorithms — silhouette plot shows cluster stability
consensus <- getConsensusMOIC(
  moic.res.list = moic.res.list,
  fig.name      = "CONSENSUS_HEATMAP",
  distance      = "euclidean",
  linkage       = "average"
)

getSilhouette(
  sil      = consensus$sil,
  fig.path = "results",
  fig.name = "SILHOUETTE",
  height   = 5.5,
  width    = 5
)

# Logit-transform betas to M-values for heatmap (improves colour scaling)
indata     <- mo.data
indata$Alu <- logit_beta(indata$Alu)
indata$L1  <- logit_beta(indata$L1)
indata$LTR <- logit_beta(indata$LTR)

# Standardise (centre + scale, truncate at ±2 SD to reduce outlier dominance)
plotdata <- getStdiz(
  data       = indata,
  halfwidth  = c(2, 2, 2),
  centerFlag = c(TRUE, TRUE, TRUE),
  scaleFlag  = c(TRUE, TRUE, TRUE)
)

# Colour palette: light grey → black → maroon (low → mid → high methylation)
Alu.col  <- c("grey90", "black", "#800000")
L1.col   <- c("grey90", "black", "#800000")
LTR.col  <- c("grey90", "black", "#800000")
col.list <- list(Alu.col, L1.col, LTR.col)

# Plot heatmap using NEMO cluster assignments as the primary subtype labels.
# Repeat this call substituting clust.res / clust.dend for each of the other 5 algorithms
# to visually compare their cluster structures.
getMoHeatmap(
  data        = plotdata,
  row.title   = c("ALU", "L1", "LTR"),
  is.binary   = c(FALSE, FALSE, FALSE),
  legend.name = c("ALU", "L1", "LTR"),
  clust.res   = moic.res.list$NEMO$clust.res,
  clust.dend  = moic.res.list$NEMO$clust.dend,
  color       = col.list,
  width       = 10,
  height      = 3,
  fig.name    = "NEMO_COMPREHENSIVE_HEATMAP"
)
```

---

## Step 6: Prepare Phenotype Data and Merge Cluster Assignments

```{r pheno-prep}
# Slim the phenotype table to clinically relevant covariates only
pheno2 <- pheno %>%
  select(submitter_id, basename, days_to_death, days_to_last_follow_up,
         days_to_recurrence, age_at_index, ajcc_pathologic_stage,
         tumor_grade, vital_status, gender, race)

# Confirm sample order matches between clustering output and phenotype table
# before merging (must be TRUE to proceed)
identical(moic.res.list$NEMO$clust.res$samID, pheno2$basename)

# Add NEMO cluster assignment to phenotype table
pheno2$NEMO <- moic.res.list$NEMO$clust.res$clust

# Assign biologically meaningful subtype names based on heatmap observations:
# Cluster 1 = Transient, 2 = Active, 3 = Repressed
pheno2$NEMO <- dplyr::recode(pheno2$NEMO,
  "1" = "Transient",
  "2" = "Active",
  "3" = "Repressed"
)
pheno2$NEMO <- factor(pheno2$NEMO, levels = c("Repressed", "Transient", "Active"))

# Binary grouping: collapse Active + Transient vs Repressed for simpler comparisons
pheno2$NEMO2 <- factor(
  ifelse(pheno2$NEMO %in% c("Transient", "Active"), "Active_Transient", "Repressed"),
  levels = c("Repressed", "Active_Transient")
)

# Remove samples with uninformative survival time (0 days) or missing stage/grade
pheno2 <- pheno2 %>%
  filter(OS != 0, !is.na(stage), !is.na(grade), grade != "GX")

# Convert survival time from days to years; rename columns for readability
pheno2$OS <- as.numeric(pheno2$days_to_last_follow_up) / 365
pheno2 <- pheno2 %>%
  rename(stage = ajcc_pathologic_stage, grade = tumor_grade,
         age = age_at_index, sex = gender)

# Encode vital status as numeric (0 = Alive, 1 = Dead) for survival models
pheno2$vital_status <- as.numeric(ifelse(pheno2$vital_status == "Alive", 0, 1))

# Standardise stage strings (replace spaces with underscores)
pheno2$stage <- gsub(" ", "_", pheno2$stage)

# Collapse race to binary (white vs others) due to small group sizes
pheno2$race <- ifelse(pheno2$race == "white", "white", "others")
```

---

## Step 7: HiTIMED Immune Deconvolution

HiTIMED estimates tumour microenvironment cell type proportions from methylation data.
Level 2 is used here to obtain three broad compartments: Tumour, Immune, and Angiogenic.

```{r hitimed}
# Run HiTIMED deconvolution for KIRC tumour samples (Level 2 = 3 compartments)
hitimedT <- HiTIMED_deconvolution(betas, tumor_type = "KIRC", h = 2, tissue_type = "tumor")
hitimedT$submitter_id <- rownames(hitimedT)

# KNN imputation for any missing deconvolution estimates
hitimedT_im      <- impute.knn(as.matrix(hitimedT), k = 10, rowmax = 0.5, colmax = 0.8)
hitimedT_im      <- as.data.frame(hitimedT_im$data)

# Add cell type proportion estimates to phenotype table
pheno2$Tumor  <- hitimedT_im$Tumor
pheno2$Immune <- hitimedT_im$Immune
pheno2$Angio  <- hitimedT_im$Angiogenic

# Create AUA risk grouping based on stage and grade
pheno2 <- pheno2 %>%
  mutate(
    AUA_risk_group = case_when(
      stage %in% c("Stage_I", "Stage_II") ~ "Low_Intermediate",
      stage %in% c("Stage_III", "Stage_IV") ~ "High",
      TRUE ~ NA_character_
    )
  )
```

---

## Step 8: Differential Methylation Analysis (DMA)

Linear models (limma) identify RE loci with significantly different methylation across subtypes,
adjusting for age, sex, stage, and immune/angiogenic cell fractions.

```{r dma}
# Assign sample identifiers as column names across all M-value matrices
colnames(Alu_M2) <- pheno$cases
colnames(L1_M2)  <- pheno$cases
colnames(LTR_M2) <- pheno$cases

# Combine all three RE family M-value matrices into one for joint modelling
RE_M <- rbind(Alu_M2, L1_M2, LTR_M2)

# Design matrix: NEMO subtype + clinical covariates
# agez, Immunez, Angioz are assumed to be z-scored continuous covariates
design_M <- model.matrix(~ 0 + NEMO + agez + sex + stage + Immunez + Angioz, data = pheno2)
colnames(design_M)[1:3] <- c("Repressed", "Transient", "Active")

fit_M <- lmFit(RE_M, design_M)

# Define pairwise contrasts between the three subtypes
contrast_M <- makeContrasts(
  TR = Transient - Repressed,   # Transient vs Repressed
  AR = Active - Repressed,      # Active vs Repressed (primary comparison)
  AT = Active - Transient,      # Active vs Transient
  levels = design_M
)

fit1_M <- eBayes(contrasts.fit(fit_M, contrast_M))

# Extract significant hits per contrast (FDR < 0.1, |logFC| > 0.2 for TR; FDR < 0.01 for AR/AT)
fit_M_TR <- topTable(fit1_M, coef = "TR", number = Inf,
                     adjust.method = "fdr", p.value = 0.1,  lfc = 0.2) %>%
  tibble::rownames_to_column("Pos")

fit_M_AR <- topTable(fit1_M, coef = "AR", number = Inf,
                     adjust.method = "fdr", p.value = 0.01, lfc = 0.2) %>%
  tibble::rownames_to_column("Pos")

fit_M_AT <- topTable(fit1_M, coef = "AT", number = Inf,
                     adjust.method = "fdr", p.value = 0.01, lfc = 0.2) %>%
  tibble::rownames_to_column("Pos")

# Combine all significant hits across contrasts (deduplicated by position)
combined_fits <- bind_rows(fit_M_TR, fit_M_AR, fit_M_AT) %>%
  distinct(Pos, .keep_all = TRUE)

# Uncomment to export:
# write.csv(combined_fits, file = p_res("EWAS_part1_combined_hits.csv"), row.names = FALSE)
```

---

## Step 9: Volcano Plot (Active vs Repressed)

```{r volcano}
# Requires EnhancedVolcano — uncomment library() above if not loaded
# volcano1 <- EnhancedVolcano(
#   fit_M_AR,
#   lab        = rownames(fit_M_AR),
#   x          = "logFC",
#   y          = "adj.P.Val",
#   col        = c("grey30", "grey30", "grey30", "maroon"),
#   title      = "Active vs Repressed",
#   pCutoff    = 0.05,
#   FCcutoff   = 1,
#   xlim       = c(-2, 2),
#   ylim       = c(0, 40)
# ) +
#   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
# volcano1
```

---

## Step 10: Extract Gene Annotations from REMP Parcels

Gene symbols and genomic coordinates are pulled from the REMP annotation parcels
for each RE family.

```{r remp-annotations-extract}
# Extract RefGene and RE position annotations from each parcel object
Alu_annot      <- as.data.frame(remparcelAlu@RefGene)
Alu_annot_REpos <- as.data.frame(remparcelAlu@RE)

L1_annot       <- as.data.frame(remparcelL1@RefGene)
L1_annot_REpos  <- as.data.frame(remparcelL1@RE)

LTR_annot      <- as.data.frame(remparcelLTR@RefGene)
LTR_annot_REpos <- as.data.frame(remparcelLTR@RE)
```

---

## Step 11: Genomic Overlap to Map RE Hits to Genes

For each RE family, narrow RE locus positions are overlapped with the broader RefGene
annotation intervals to assign gene symbols and chromosomal context to each hit.

```{r genomic-overlap}
# --- Alu ---
gr_alu <- GRanges(
  seqnames = Alu_annot$seqnames,
  ranges   = IRanges(start = Alu_annot$start, end = Alu_annot$end),
  strand   = if ("strand" %in% names(Alu_annot)) Alu_annot$strand else "*"
)
mcols(gr_alu) <- Alu_annot %>% select(-seqnames, -start, -end, -strand)

gr_pos <- GRanges(
  seqnames = Alu_annot_REpos$seqnames,
  ranges   = IRanges(start = Alu_annot_REpos$start, end = Alu_annot_REpos$end),
  strand   = if ("strand" %in% names(Alu_annot_REpos)) Alu_annot_REpos$strand else "*"
)
mcols(gr_pos) <- Alu_annot_REpos %>% select(-seqnames, -start, -end, -strand)

ov <- findOverlaps(gr_pos, gr_alu, ignore.strand = TRUE)
Alu_ewas_annot <- bind_cols(
  as.data.frame(gr_pos[queryHits(ov)]),
  as.data.frame(mcols(gr_alu)[subjectHits(ov), ])
)

# --- L1 ---
gr_L1 <- GRanges(
  seqnames = L1_annot$seqnames,
  ranges   = IRanges(start = L1_annot$start, end = L1_annot$end),
  strand   = if ("strand" %in% names(L1_annot)) L1_annot$strand else "*"
)
mcols(gr_L1) <- L1_annot %>% select(-seqnames, -start, -end, -strand)

gr_pos <- GRanges(
  seqnames = L1_annot_REpos$seqnames,
  ranges   = IRanges(start = L1_annot_REpos$start, end = L1_annot_REpos$end),
  strand   = if ("strand" %in% names(L1_annot_REpos)) L1_annot_REpos$strand else "*"
)
mcols(gr_pos) <- L1_annot_REpos %>% select(-seqnames, -start, -end, -strand)

ov <- findOverlaps(gr_pos, gr_L1, ignore.strand = TRUE)
L1_ewas_annot <- bind_cols(
  as.data.frame(gr_pos[queryHits(ov)]),
  as.data.frame(mcols(gr_L1)[subjectHits(ov), ])
)

# --- LTR ---
gr_LTR <- GRanges(
  seqnames = LTR_annot$seqnames,
  ranges   = IRanges(start = LTR_annot$start, end = LTR_annot$end),
  strand   = if ("strand" %in% names(LTR_annot)) LTR_annot$strand else "*"
)
mcols(gr_LTR) <- LTR_annot %>% select(-seqnames, -start, -end, -strand)

gr_pos <- GRanges(
  seqnames = LTR_annot_REpos$seqnames,
  ranges   = IRanges(start = LTR_annot_REpos$start, end = LTR_annot_REpos$end),
  strand   = if ("strand" %in% names(LTR_annot_REpos)) LTR_annot_REpos$strand else "*"
)
mcols(gr_pos) <- LTR_annot_REpos %>% select(-seqnames, -start, -end, -strand)

ov <- findOverlaps(gr_pos, gr_LTR, ignore.strand = TRUE)
LTR_ewas_annot <- bind_cols(
  as.data.frame(gr_pos[queryHits(ov)]),
  as.data.frame(mcols(gr_LTR)[subjectHits(ov), ])
)
```

---

## Step 12: Annotate DMA Results

Significant DMA hits are joined with genomic annotations to produce interpretable output
tables with gene symbols, RE family/class, and chromosomal positions.

```{r annotate-dma-results}
# Subset each annotation table to positions present in the combined significant hits
Alu_res  <- Alu_ewas_annot[Alu_ewas_annot$Index %in% combined_fits$Pos, ]
L1_res   <- L1_ewas_annot[L1_ewas_annot$Index   %in% combined_fits$Pos, ]
LTR_res  <- LTR_ewas_annot[LTR_ewas_annot$Index %in% combined_fits$Pos, ]
Res_annot <- rbind(Alu_res, L1_res, LTR_res)

# Active vs Repressed annotated results (primary comparison)
Alu_ARres  <- Alu_ewas_annot[Alu_ewas_annot$Index %in% fit_M_AR$Pos, ]
L1_ARres   <- L1_ewas_annot[L1_ewas_annot$Index   %in% fit_M_AR$Pos, ]
LTR_ARres  <- LTR_ewas_annot[LTR_ewas_annot$Index %in% fit_M_AR$Pos, ]
Res_ARannot <- rbind(Alu_ARres, L1_ARres, LTR_ARres)

# Merge with logFC and p-value from DMA
Res_ARannot <- merge(
  Res_ARannot,
  fit_M_AR[, c("Pos", "logFC", "P.Value")],
  by.x = "Index", by.y = "Pos"
)

# Select key columns and remove duplicate genomic positions
Res_ARannot2 <- Res_ARannot %>%
  dplyr::select(Index, seqnames, start, end, repClass, repFamily, GeneSymbol, logFC, P.Value) %>%
  filter(!duplicated(start))

Res_ARannot2

# Uncomment to export:
# write.csv(Res_ARannot2, file = p_res("Res_ARannot2_ActiveVsRepressed.csv"), row.names = FALSE)
```
