---
title: "Survival RE Analysis"
author: "Chinaza Nnam"
date: "2026-02-27"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

## Overview

This script performs survival analysis on TCGA-KIRC samples stratified by RE methylation
subtypes defined in `01_TCGA_REMP_RE_Analysis.Rmd`.

It covers:

1. Kaplan–Meier survival curves by NEMO subtype
2. Median survival extraction per group
3. Multivariable Cox proportional hazards models (AJCC stage and AUA risk groupings)
4. Sensitivity analysis including race as a covariate
5. Forest plot visualising hazard ratios across models

**Required input objects (carried over from `01_TCGA_REMP_RE_Analysis.Rmd`):**
- `pheno2` — phenotype table with columns: `OS`, `vital_status`, `NEMO`, `age`, `sex`,
  `stage`, `AUA_risk_group`, `Immune`, `Angio`, `race`

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "results/figures/"
)
```

---

## Load Packages

```{r packages}
suppressPackageStartupMessages({
  library(survival)   # Surv(), survfit(), coxph()
  library(survminer)  # ggsurvplot()
  library(dplyr)      # data wrangling
  library(tibble)     # rownames_to_column()
  library(broom)      # tidy model outputs
  library(ggplot2)    # forest plot
})
```

---

## Step 1: Kaplan–Meier Survival Analysis

Fit KM curves stratified by NEMO subtype and plot with p-value and risk table.
Colours: green = Repressed, amber = Transient, red = Active.

```{r km-curve}
# Fit KM curves for each NEMO subtype using overall survival (OS, years)
fit_km <- survfit(Surv(OS, vital_status) ~ NEMO, data = pheno2)

ggsurvplot(
  fit_km,
  data         = pheno2,
  pval         = TRUE,           # log-rank p-value
  risk.table   = TRUE,           # number at risk below plot
  palette      = c("#117733", "#fda000", "#ae2029"),  # Repressed, Transient, Active
  legend.title = "RE Subtype",
  xlab         = "Time (Years)",
  ylab         = "Survival Probability",
  title        = "Overall Survival by RE Methylation Subtype (TCGA-KIRC, k=3)"
)
```

---

## Step 2: Median Survival per Group

```{r median-survival}
# Extract median survival time and 95% CI for each NEMO group
median_surv <- summary(fit_km)$table %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Group") %>%
  transmute(
    NEMO_Group            = sub("^NEMO=", "", Group),
    Median_Survival_years = median,
    Lower_95CI            = `0.95LCL`,
    Upper_95CI            = `0.95UCL`
  )

median_surv
```

---

## Step 3: Multivariable Cox Proportional Hazards Models

Four models are tested using two staging systems (AJCC and AUA risk group), each with and
without immune/angiogenic cell fraction adjustment from HiTIMED:

- **Model 1A:** NEMO + age + sex + AJCC stage
- **Model 1B:** Model 1A + scaled Immune + Angio fractions
- **Model 2A:** NEMO + age + sex + AUA risk group
- **Model 2B:** Model 2A + scaled age + scaled Immune + Angio fractions

```{r cox-models}
# Model 1A: AJCC stage adjustment only
fit_Model1A <- coxph(Surv(OS, vital_status) ~ NEMO + age + sex + stage,
                     data = pheno2)

# Model 1B: AJCC stage + immune microenvironment adjustment
fit_Model1B <- coxph(Surv(OS, vital_status) ~ NEMO + age + sex + stage +
                       scale(Immune) + scale(Angio),
                     data = pheno2)

# Model 2A: AUA risk group adjustment only
fit_Model2A <- coxph(Surv(OS, vital_status) ~ NEMO + age + sex + AUA_risk_group,
                     data = pheno2)

# Model 2B: AUA risk group + scaled covariates + immune microenvironment adjustment
fit_Model2B <- coxph(Surv(OS, vital_status) ~ NEMO + scale(age) + sex + AUA_risk_group +
                       scale(Immune) + scale(Angio),
                     data = pheno2)

# Print summaries
summary(fit_Model1A)
summary(fit_Model1B)
summary(fit_Model2A)
summary(fit_Model2B)
```

---

## Step 4: Sensitivity Analysis — Race Inclusion

Race was tested as an additional covariate. It did not significantly alter model estimates
and was excluded from primary models to preserve statistical power.

```{r sensitivity-race}
# Sensitivity model: AUA risk group + immune fractions + race
fit_race <- coxph(Surv(OS, vital_status) ~ NEMO + scale(age) + sex + AUA_risk_group +
                    scale(Immune) + scale(Angio) + race,
                  data = pheno2)

summary(fit_race)
```

---

## Step 5: Extract Hazard Ratios for Active vs Repressed (Reference)

A helper function extracts HR, 95% CI, and p-value for the Active NEMO contrast from any
Cox model. Repressed is the reference level.

```{r extract-hr-function}
# Helper: extracts HR, CI, and p-value for the Active NEMO contrast
# model       — a fitted coxph object
# nemo_term   — prefix used to identify NEMO rows in the coefficient table
extract_nemo_hr <- function(model, nemo_term = "NEMO") {
  sm        <- summary(model)
  nemo_rows <- grep(paste0("^", nemo_term), rownames(sm$coefficients))
  active_row <- grep("Active", rownames(sm$coefficients)[nemo_rows], value = TRUE)

  if (length(active_row) == 0) stop("Active level not found in model coefficients.")

  coef_row <- sm$coefficients[active_row, ]
  ci_row   <- sm$conf.int[active_row, ]

  data.frame(
    term  = active_row,
    HR    = ci_row["exp(coef)"],
    lower = ci_row["lower .95"],
    upper = ci_row["upper .95"],
    pval  = coef_row["Pr(>|z|)"],
    row.names = NULL
  )
}

# Extract HRs from each model
res1 <- extract_nemo_hr(fit_Model1A)
res2 <- extract_nemo_hr(fit_Model1B)
res3 <- extract_nemo_hr(fit_Model2A)
res4 <- extract_nemo_hr(fit_Model2B)

# Combine into a single data frame for plotting
forest_df <- rbind(
  cbind(Model = "Model 1A", res1),
  cbind(Model = "Model 1B", res2),
  cbind(Model = "Model 2A", res3),
  cbind(Model = "Model 2B", res4)
)

# Set factor levels so models display top-to-bottom in correct order on the forest plot
forest_df$Model <- factor(forest_df$Model,
                          levels = rev(c("Model 1A", "Model 1B", "Model 2A", "Model 2B")))

forest_df
```

---

## Step 6: Forest Plot of Hazard Ratios Across Models

Each point = HR for Active vs Repressed; bars = 95% CI.
Dashed red line at HR = 1 (no effect). All models adjust for at least age, sex, and a
staging variable.

```{r forest-plot}
ggplot(forest_df, aes(x = Model, y = HR, ymin = lower, ymax = upper)) +
  geom_pointrange(size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Hazard Ratios for Active RE Subtype Across Models\n(Reference: Repressed)",
    y     = "Hazard Ratio (HR)",
    x     = ""
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text  = element_text(color = "black")
  )

# Uncomment to save:
# ggsave(filename = "results/figures/forest_plot_HR.pdf", width = 7, height = 4)
```
