---
title: "Hypoxia and Immune Response Signatures — RE Analysis"
author: "Chinaza Nnam"
date: "2026-02-27"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

## Overview

This script characterises the tumour microenvironment and immune/hypoxia gene expression
signatures across RE methylation subtypes (NEMO clusters) defined in
`01_TCGA_REMP_RE_Analysis.Rmd`.

It covers:

1. HiTIMED deconvolution at 12-cell-type resolution (h = 6)
2. Alignment of TCGA-KIRC methylation and RNA-seq (TPM) samples
3. Extraction of hypoxia (EPAS1, HIF1A) and immune checkpoint (PDCD1) gene expression
4. ssGSEA scoring for cGAS-STING, interferon-stimulated genes (ISG), and T cell
   exhaustion (terminal and progenitor) signatures
5. Violin/boxplots comparing signatures across RE subtypes

**Required input objects (carried over from `01_TCGA_REMP_RE_Analysis.Rmd`):**
- `betas` — CpG beta matrix used for REMP predictions to be used for HiTIMED too
- `pheno2` — phenotype table with NEMO assignments and clinical covariates
- `tcga_tpm` — gene × sample TPM matrix for TCGA-KIRC RNA-seq
- `pheno_methrna` — phenotype table pre-filtered to samples with both methylation and RNA-seq data

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "results/figures/"
)
```

---

## Load Packages

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)         # data wrangling
  library(tidyr)         # reshaping
  library(ggplot2)       # base plotting
  library(ggpubr)        # geom_pwc() pairwise comparisons
  library(survival)      # survival objects
  library(survminer)     # KM plots
  library(GSVA)          # ssGSEA scoring
  library(BiocParallel)  # parallel backend for GSVA
  library(tibble)        # rownames_to_column()
})
```

---

## Step 1: HiTIMED Deconvolution at 12-Cell-Type Resolution

Running HiTIMED at hierarchy level 6 (h = 6) returns 12 immune/stromal cell type proportions.
These are joined onto `pheno2` to create `pheno_hi` for downstream cell-type comparisons.

```{r hitimed-12ct}
# Run HiTIMED at h=6 for 12-cell-type resolution on TCGA-KIRC tumour samples
hitimed12 <- HiTIMED_deconvolution(betas, tumor_type = "KIRC", h = 6, tissue_type = "tumor")

# Use rownames (sample IDs) as a join key
hitimed12$submitter_id <- rownames(hitimed12)

# Merge 12-cell-type proportions with the primary phenotype table
pheno_hi <- left_join(pheno2, hitimed12, by = "submitter_id")
```

---

## Step 2: Align Methylation and RNA-seq Samples

Only samples with both methylation and RNA-seq data are retained for gene expression analyses.
`pheno_methrna` must already exist as a pre-filtered phenotype table; this step confirms
alignment and adds HiTIMED cell type proportions.

```{r align-methrna}
# Identify samples present in both the phenotype table and the TPM matrix
common_ids <- intersect(pheno_methrna$submitter_id, colnames(tcga_tpm))

# Subset TPM matrix and phenotype table to shared samples only
tpm_methrna  <- tcga_tpm[, common_ids]
pheno_methrna <- pheno2[match(common_ids, pheno2$submitter_id), ]

# Confirm column order in TPM matches row order in phenotype (required for downstream models)
stopifnot(all(colnames(tpm_methrna) == pheno_methrna$submitter_id))

# Add 12-cell-type HiTIMED proportions to the aligned phenotype table
pheno_methrna <- left_join(pheno_methrna, hitimed12, by = "submitter_id")
```

---

## Step 3: Extract Hypoxia and Checkpoint Gene Expression

EPAS1 (HIF-2α), HIF1A (HIF-1α), and PDCD1 (PD-1) TPM values are pulled directly from the
aligned TPM matrix and added as columns to `pheno_methrna` for boxplot comparisons.

```{r extract-gene-expression}
# Genes of interest: hypoxia transcription factors and immune checkpoint marker
hifs_pd1 <- c("EPAS1", "HIF1A", "PDCD1")

# Confirm all genes are present in the TPM matrix before subsetting
missing_genes <- setdiff(hifs_pd1, rownames(tpm_methrna))
if (length(missing_genes) > 0) {
  warning("The following genes were not found in tpm_methrna: ",
          paste(missing_genes, collapse = ", "))
}

# Transpose the gene subset so samples are rows, then join to phenotype by submitter_id
tpm_genes <- as.data.frame(t(tpm_methrna[intersect(hifs_pd1, rownames(tpm_methrna)), ,
                                          drop = FALSE]))
tpm_genes$submitter_id <- rownames(tpm_genes)

pheno_methrna <- left_join(pheno_methrna, tpm_genes, by = "submitter_id")
```

---

## Step 4: ssGSEA Scoring

Single-sample GSEA (ssGSEA) scores are computed for four immune/hypoxia gene sets.
Each score reflects the relative enrichment of that pathway in each individual sample.

**Gene sets:**
- **cGAS-STING** — innate DNA-sensing pathway
- **ISG** — interferon-stimulated gene response
- **Terminal exhaustion** — dysfunctional T cell markers (TOX, PDCD1, HAVCR2, etc.)
- **Progenitor exhaustion** — stem-like exhausted T cell markers (TCF7, CCR7, etc.)

```{r define-gene-sets}
# cGAS-STING innate immune sensing pathway
cgas <- c("CGAS", "STING1", "TBK1", "IKBKE", "IRF3", "POLG", "TREX1", "ENPP1")

# Interferon-stimulated genes (ISG) — broad antiviral transcriptional response
isg <- c(
  "IFIT1", "IFIT2", "IFIT3",
  "ISG15", "ISG20",
  "MX1", "MX2",
  "OAS1", "OAS2", "OAS3", "OASL",
  "IRF7", "IRF9",
  "STAT1", "STAT2",
  "IFIH1", "IFI27", "IFI35", "IFI44", "IFI44L", "IFI6",
  "RSAD2", "BST2",
  "CXCL10", "CXCL11",
  "DDX58", "DHX58",
  "HERC5", "PLSCR1", "USP18",
  "GBP1", "GBP2", "GBP4", "GBP5",
  "SECTM1"
)

# Terminal T cell exhaustion markers
# Note: NR4A1-3 are written as explicit symbols (avoiding en-dash encoding issues)
# ENTPD1 = CD39
term_ex <- c("TOX", "EOMES", "PRDM1", "BATF",
             "NR4A1", "NR4A2", "NR4A3",
             "PDCD1", "HAVCR2", "TIGIT", "CTLA4", "LAG3", "ENTPD1")

# Progenitor / stem-like exhausted T cell markers
prog_ex <- c("TCF7", "CCR7", "CXCR5", "IL7R")
```

```{r ssgsea-helper}
# Helper function: runs ssGSEA for a single gene set against the expression matrix
# expr_mat  — genes x samples matrix (TPM)
# genes     — character vector of gene symbols for the gene set
# set_name  — label used in warning messages
run_ssgsea_one <- function(expr_mat, genes, set_name) {
  genes_use <- intersect(genes, rownames(expr_mat))

  if (length(genes_use) == 0) {
    message("No genes found in expression matrix for set: ", set_name)
    return(rep(NA_real_, ncol(expr_mat)))
  }
  if (length(genes_use) < length(genes)) {
    message("Set '", set_name, "' is missing: ",
            paste(setdiff(genes, genes_use), collapse = ", "))
  }

  # Build ssGSEA parameter object and run — alpha = 0.25 is the standard ssGSEA weighting
  param <- ssgseaParam(
    expr_mat,
    list(tmp = genes_use),
    assay      = NA_character_,
    annotation = NULL,
    minSize    = 1,
    maxSize    = Inf,
    alpha      = 0.25,
    normalize  = TRUE
  )

  as.numeric(GSVA::gsva(param, BPPARAM = BiocParallel::SerialParam(progressbar = TRUE)))
}
```

```{r run-ssgsea}
# Compute ssGSEA scores for all four gene sets
cgas_scores   <- run_ssgsea_one(tpm_methrna, cgas,    "cgas_sting")
isg_scores    <- run_ssgsea_one(tpm_methrna, isg,     "isg")
termex_scores <- run_ssgsea_one(tpm_methrna, term_ex, "terminal_exhaustion")
progex_scores <- run_ssgsea_one(tpm_methrna, prog_ex, "progenitor_exhaustion")

# Assemble scores into a data frame aligned to the TPM column order
scores_df <- data.frame(
  submitter_id = colnames(tpm_methrna),
  cgas_sting   = cgas_scores,
  isg          = isg_scores,
  term_ex      = termex_scores,
  prog_ex      = progex_scores,
  stringsAsFactors = FALSE
)

# Join ssGSEA scores onto the aligned phenotype table
pheno_methrna <- left_join(pheno_methrna, scores_df, by = "submitter_id")
```

---

## Step 5: Violin / Boxplots by RE Subtype

Each plot compares a signature or gene expression value across NEMO subtypes.
Pairwise BH-adjusted p-values are added with `geom_pwc()`.

To compare a different variable, change the `y` aesthetic and `labs(y = ...)` accordingly.
Available variables in `pheno_methrna`: `EPAS1`, `HIF1A`, `PDCD1`, `cgas_sting`, `isg`,
`term_ex`, `prog_ex`, and any HiTIMED cell type column.

```{r plot-helper}
# Reusable function to generate violin + boxplot for any numeric variable
# df       — data frame (pheno_methrna)
# y_var    — column name as a string (e.g. "EPAS1")
# y_label  — y-axis label for the plot
plot_by_nemo <- function(df, y_var, y_label) {
  ggplot(df, aes(x = NEMO, y = as.numeric(.data[[y_var]]), fill = NEMO)) +
    geom_violin(trim = FALSE, alpha = 0.4, color = NA) +
    geom_boxplot(width = 0.20, alpha = 0.9, outlier.shape = NA, color = "black") +
    scale_fill_manual(values = c(
      "Repressed" = "#117733",
      "Transient" = "#fdb813",
      "Active"    = "#ae2029"
    )) +
    geom_pwc(
      method          = "wilcox.test",
      p.adjust.method = "BH",
      label           = "p.adj.signif",
      label.size      = 5
    ) +
    labs(y = y_label, x = NULL) +
    theme_bw(base_size = 20) +
    theme(
      legend.position  = "none",
      axis.title       = element_text(face = "bold", size = 25),
      axis.text.x      = element_text(face = "bold", size = 20),
      axis.text.y      = element_text(size = 20),
      plot.title       = element_text(hjust = 0.5, face = "bold", size = 18),
      panel.grid       = element_blank()
    )
}
```

```{r plots}
# Hypoxia transcription factors
plot_by_nemo(pheno_methrna, "EPAS1",  "EPAS1 / HIF-2α (TPM)")
plot_by_nemo(pheno_methrna, "HIF1A",  "HIF1A / HIF-1α (TPM)")

# Immune checkpoint
plot_by_nemo(pheno_methrna, "PDCD1",  "PDCD1 / PD-1 (TPM)")

# Immune response pathway scores
plot_by_nemo(pheno_methrna, "cgas_sting", "cGAS-STING Score (ssGSEA)")
plot_by_nemo(pheno_methrna, "isg",        "ISG Score (ssGSEA)")
plot_by_nemo(pheno_methrna, "term_ex",    "Terminal Exhaustion Score (ssGSEA)")
plot_by_nemo(pheno_methrna, "prog_ex",    "Progenitor Exhaustion Score (ssGSEA)")

# Uncomment to save individual plots, e.g.:
# ggsave("results/figures/EPAS1_by_NEMO.pdf", width = 6, height = 6)
```
