---
title: "Validation Cohort Analysis"
author: "CN"
date: "2026-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This script performs multi-omics integrative clustering (MOVICS/NEMO) on CPTAC validation cohort 
DNA methylation data for repetitive element (RE) families: **Alu**, **L1**, and **LTR**.

**Prerequisites:** REMP predictions and QC must be run on the CPTAC beta matrix prior to this 
script, following the same pipeline as in `TCGA_REMP_RE_Analysis`. The CPTAC beta matrix should 
be subsetted to CpGs present in the 450k array landscape. All CPTAC-derived objects are named 
with the suffix `_cp`.

**Downstream:** After clustering, merge results with phenotype data and proceed to survival and 
immune/hypoxia signature analyses as in:
- `Survival_RE_Analysis`
- `Hypoxia_ImmuneResponse_Signatures_Analysis`

---

## Step 1: Identify Common CpGs Between TCGA and CPTAC

Combine the top 1,000 elite CpGs from each RE family in the TCGA results, then subset the 
corresponding CPTAC beta matrices to retain only those CpGs. This ensures the validation cohort 
uses the same features identified in the discovery cohort.

```{r common-cpgs}
# Combine top 1k elite CpGs across all three RE families from TCGA discovery cohort
RE_tcga <- rbind(Alu_1k$elite.dat, L1_1k$elite.dat, LTR_1k$elite.dat)

# Subset each CPTAC RE beta matrix to retain only CpGs present in the TCGA top 1k set
common_Alu <- Alu_beta2[rownames(Alu_beta2) %in% rownames(RE_tcga), ]
common_L1  <- L1_beta2[rownames(L1_beta2)   %in% rownames(RE_tcga), ]
common_LTR <- LTR_beta2[rownames(LTR_beta2) %in% rownames(RE_tcga), ]
```

---

## Step 2: Multi-Omics Integrative Clustering (NEMO)

Assemble the multi-omics data list and run NEMO clustering with k = 3, matching the cluster 
number used in the TCGA discovery cohort. `getClustNum()` is commented out as the optimal 
cluster number has already been determined.

```{r nemo-clustering}
# Assemble multi-omics input list for MOVICS
mo.data_valid <- list(Alu = common_Alu, L1 = common_L1, LTR = common_LTR)

# NOTE: Optimal cluster number (k=3) was determined in the TCGA discovery cohort.
# Uncomment below to re-evaluate if needed:
# optk.RE <- getClustNum(data        = mo.data_valid,
#                        is.binary   = c(F, F, F),
#                        try.N.clust = 2:8)

# Run NEMO clustering with k=3; all three data types are continuous (gaussian)
moic.res.list_valid <- getMOIC(data        = mo.data_valid,
                               methodslist = list("NEMO"),
                               N.clust     = 3,
                               type        = c("gaussian", "gaussian", "gaussian"))
```

---

## Step 3: Logit Transform Beta Values and Normalise for Heatmap

Beta values (range 0–1) are logit-transformed to M-values (range −∞ to +∞) prior to 
standardisation, which improves statistical properties for visualisation. The standardised 
data are then used to generate a multi-omics heatmap.

```{r logit-transform-and-heatmap}
# Copy the multi-omics list to avoid modifying the original beta matrices
indata_valid <- mo.data_valid

# Logit-transform beta values to M-values: log2(beta / (1 - beta))
# This linearises the scale and is preferred for methylation visualisation
indata_valid$Alu <- log2(indata_valid$Alu / (1 - indata_valid$Alu))
indata_valid$L1  <- log2(indata_valid$L1  / (1 - indata_valid$L1))
indata_valid$LTR <- log2(indata_valid$LTR / (1 - indata_valid$LTR))

# Standardise (centre and scale) each data type for heatmap visualisation
# halfwidth = 2 truncates extreme values at ±2 SD to reduce colour saturation from outliers
plotdata_valid <- getStdiz(data       = indata_valid,
                           halfwidth  = c(2, 2, 2),
                           centerFlag = c(T, T, T),
                           scaleFlag  = c(T, T, T))

# Generate multi-omics heatmap annotated with NEMO cluster assignments
getMoHeatmap(data        = plotdata_valid,
             row.title   = c("ALU", "L1", "LTR"),
             is.binary   = c(F, F, F),
             legend.name = c("ALU", "L1", "LTR"),
             clust.res   = moic.res.list_valid$NEMO$clust.res,  # cluster membership
             clust.dend  = moic.res.list_valid$NEMO$clust.dend, # dendrogram
             color       = col.list)
```

---

## Next Steps

Merge `moic.res.list_valid$NEMO$clust.res` with the CPTAC phenotype (`pheno`) data and 
proceed to downstream analyses:

- **`Survival_RE_Analysis`** — Kaplan–Meier and Cox regression by cluster
- **`Hypoxia_ImmuneResponse_Signatures_Analysis`** — Signature scoring by cluster
